Letter: Performance analysis of \\ high-resolution ice sheet simulations

### 2022-06-26 paper/perfmod.tex
apparently completed performance analysis section

\begin{table*}[ht]
{\normalsize
\begin{tabular}{cll}
\emph{name} & \emph{meaning} & \emph{units} \\ \hline
$\alpha$    & one fixed-geometry Stokes velocity solution requires $O(n^{1+\alpha})${\large \strut} work\\
$\beta$     & one implicit SIA geometry-update solution requires $O(n^{1+\beta})$ work \\
$\gamma$    & one implicit Stokes geometry-update and velocity solution requires $O(n^{1+\gamma})$ work \\
$D$         & representative geometric (SIA) diffusivity of an ice sheet & $\text{km}^2 \text{a}^{-1}$ \\
$L$         & width of simulation domain & km \\
$m$         & degrees of freedom: number of nodes in the horizontal mesh \\
$q$         & time steps per model year needed to resolve modeled climate interactions & $\text{a}^{-1}$ \\
$\Delta t$  & length of time step in model years & a \\
$U$         & representative horizontal ice velocity & $\text{km}\,\text{a}^{-1}$ \\
$\Delta x$  & representative width (diameter) of map-plane mesh cells & km
\end{tabular}
}
\caption{Parameters for performance analysis; $\alpha,\beta,\gamma,n$ are pure numbers.}
\label{tab:notation}
\end{table*}

@@

A numerical ice sheet model uses $m$ ice thickness or surface elevation variables, one \emph{degree of freedom} per mesh node.  It requires $O(m)$ memory to store the instantaneous model state, including thermodynamical variables for example, if we assume that the vertical mesh/grid has an \emph{a priori} bounded resolution.\footnote{This assumption reflects common usage \citep[for example]{Aschwandenetal2019,Brinkerhoffetal2017,Hoffmanetal2018}.  However, the \cite{IsaacStadlerGhattas2015} Stokes and \cite{BrownSmithAhmadia2013} higher-order multigrid solvers use 3D refinement, though with certain modifications regarding vertical refinement.}  The amount of fast memory needed by a running ice sheet simulation is also $O(m)$ if prior states are discarded or transferred to storage.  Ice sheet models also have $O(m)$ velocity variables, but these are not state variables because a very-viscous stress balance will compute velocity as a function of the true state variables.

@@

For each explicit time step the computational cost of a velocity (or velocity/pressure) solution of the stress balance equations is determined by solver design.  For the Stokes model we suppose one such solution requires $O(m^{1+\alpha})$ floating point operations.\footnote{Many considerations are suppresed in our simple analysis, including memory latency, memory bandwidth, and process/thread/GPU parallelism.}

The power $\alpha$ depends on the design of the stress balance solver.  A direct linear algebra Stokes solver implementation might yield $\alpha=3$, or $\alpha \approx 2$ when exploiting sparsity, but a multigrid method, such as the algebraic multigrid \citep{Trottenbergetal2001} solver by \cite{IsaacStadlerGhattas2015}, reduces $\alpha$.  Their Antarctic ice sheet results show Newton-Krylov iterations \citep{Bueler2021} growing only slowly under mesh refinement, suggesting perhaps $\alpha\approx 0.2$ \citep[see Table 8.1]{IsaacStadlerGhattas2015}.
% see data/scalingisaac.m
A geometric multigrid method applied to the higher-order stress balance equations \citep{BrownSmithAhmadia2013} suggests $\alpha$ also close to zero for certain simplified geometries.\footnote{The \cite{IsaacStadlerGhattas2015} and \cite{BrownSmithAhmadia2013} results also show good parallel scaling, something not considered in our simplified analysis.}  At the even easier end, the isothermal SIA velocity computation is a trivialization of the Stokes problem in which the vertically-averaged velocity at each node is computed by a pointwise formula, $O(1)$ work; a velocity solution requires only $O(m)$ work ($\alpha=0$).

When analyzing solver scaling in this simple manner, one must be aware that the constant in $O(m^{1+\alpha})$ can be very large.  Furthermore it will vary greatly based on solver design.

Regardless of the stress balance source, an explicit time-stepping scheme applies the mass continuity equation to update the ice thickness using $O(m)$ work.  That is, once the velocity is computed for the geometry from the previous time step, old thickness values are replaced by new ones by a formula needing only $O(1)$ work per mesh node.

\newcommand{\oo}[1]{\displaystyle O\left(#1\right)}
\setlength{\tabcolsep}{5pt}
\renewcommand{\arraystretch}{1.5}
\begin{table*}[ht]
{\normalsize
\begin{tabular}{llll}
\emph{time-stepping} & \emph{dynamics} & \emph{flops per model year} & \emph{pessimistic} \\ \hline
explicit & SIA    & $\oo{\frac{D\, L^2}{\Delta x^4}} = \oo{\frac{D\, m^2}{L^2}}${\Huge \strut} \\
explicit & Stokes & $\oo{\frac{U L^{2+2\alpha}}{\Delta x^{3+2\alpha}}} = \oo{\frac{U m^{1.5+\alpha}}{L}}${\Huge \strut}\phantom{x} & $\oo{\frac{D\, L^{2+2\alpha}}{\Delta x^{4+2\alpha}}} = \oo{\frac{D\,m^{2+\alpha}}{L^2}}$ \\
implicit & SIA    & $\oo{\frac{q\, L^{2+2\beta}}{\Delta x^{2+2\beta}}} = \oo{q\, m^{1+\beta}}${\Huge \strut} \\
implicit & Stokes & $\oo{\frac{q\, L^{2+2\gamma}}{\Delta x^{2+2\gamma}}} = \oo{q\, m^{1+\gamma}}${\Huge \strut}
\end{tabular}
}
\caption{Asymptotic scaling of computational work, measured by floating point operations per model year, for map-plane (2D) time-stepping numerical ice sheet simulations, in the high resolution limit where $\Delta x\to 0$ and $m\to\infty$.  See Table \ref{tab:notation} for notation.}
\label{tab:performancemodel}
\end{table*}

Now suppose that a numerical model takes time steps of $\Delta t$ model years, equivalently $\Delta t^{-1}$ time steps per model year.  Recalling the scaling of $\Delta x$ with $m$, namely that $\Delta x = O(L m^{-1/2})$, equivalently $m=O(L^2 \Delta x^{-2})$, for explicit schemes we may write the stability restriction as a required number of time-steps per model year in order to maintain stability, namely
\begin{equation}
\frac{1}{\Delta t} > \oo{\frac{D}{\Delta x^2}} = \oo{\frac{D m}{L^2}} \label{eq:explicitsiarequired}
\end{equation}
in the SIA and pessimistic-Stokes cases.  In the optimistic-Stokes case the estimate becomes
\begin{equation}
\frac{1}{\Delta t} > \oo{\frac{U}{\Delta x}} = \oo{\frac{U m^{1/2}}{L}}. \label{eq:explicitoptstokesrequired}
\end{equation}

The number of time steps per model year is multiplied by the per-step computational cost to give a work estimate for each model year in a simulation.  The results so far are shown in the ``explicit'' rows of Table \ref{tab:performancemodel}.

For unconditionally-stable implicit methods, however, $\Delta t$ does not scale with grid resolution; it has fixed value $\Delta t = 1/q$ determined by the need to resolve climatic interactions.  On the other hand, the per-step expense is much greater because nontrivial equations, indeed a free-boundary problem, must be solved.  FIXME for the unconditionally-stable implicit geometric-update scheme in the SIA $O(m^{1+\beta})$ work; \cite{Bueler2016} simplified dome geometry case has $\beta=1.8$, which is not adequate; it is not multigrid
% see data/bueler.perf and data/scalingbueler.m
the steady-state SIA solver in \cite{Bueler2016} allows $\Delta t=+\infty$, with $\Delta t = 100 \,\text{a}$ as a recovery strategy in the case of solver non-convergence

For an implicit Stokes simulation the geometry/velocity solve is assumed to be, in the absense of constraining research, $O(m^{1+\gamma})$ for some $\gamma>0$ to be determined.  The nMCD method proposed in [Bueler 2022] is expensive but multigrid, so it may turn out that $\gamma$ is small.

FIXME emphasize the estimate reported in Table \ref{tab:performancemodel} must be multiplied by a potentially very large scheme-dependent constant.

@@

FIXME The real promise of multigrid methods for the implicit, coupled geometry-update plus velocity solution problems is to make $\beta,\gamma$ seriously smaller.  Specifically $\gamma < 1$ might be regarded as the goal of the next few decades of work on Stokes models.  However, this aspiration is long-term because work is barely started on it.

>>>>

\begin{table*}[ht]
{\normalsize
\begin{tabular}{cll}
\emph{name} & \emph{meaning} & \emph{units} \\ \hline
$\alpha$    & one fixed-geometry Stokes velocity solution requires $O(n^{1+\alpha})${\large \strut} work\\
$\beta$     & one implicit SIA geometry-update (and velocity) solution requires $O(n^{1+\beta})$ work \\
$\gamma$    & one implicit, coupled Stokes geometry-update and velocity solution requires $O(n^{1+\gamma})$ work \\
$D$         & representative geometric (SIA) diffusivity of an ice sheet & $\text{km}^2 \text{a}^{-1}$ \\
$L$         & width of simulation domain & km \\
$m$         & degrees of freedom: number of nodes in the horizontal mesh \\
$q$         & time steps per model year needed to resolve modeled climate interactions & $\text{a}^{-1}$ \\
$\Delta t$  & length of time step in model years & a \\
$U$         & representative horizontal ice velocity & $\text{km}\,\text{a}^{-1}$ \\
$\Delta x$  & representative width (diameter) of map-plane mesh cells & km
\end{tabular}
}
\caption{Parameters for performance analysis; $\alpha,\beta,\gamma,n$ are pure numbers.}
\label{tab:notation}
\end{table*}

@@

A numerical ice sheet model uses $m$ ice thickness or surface elevation variables, one \emph{degree of freedom} per mesh node.  It requires $O(m)$ memory to store the instantaneous model state, including thermodynamical variables for example, if we assume that the vertical mesh/grid has an \emph{a priori} bounded resolution.\footnote{This assumption reflects common usage \citep[for example]{Aschwandenetal2019,Brinkerhoffetal2017,Hoffmanetal2018,Lengetal2012}.  However, the \cite{BrownSmithAhmadia2013,IsaacStadlerGhattas2015} models use 3D refinement, though with certain modifications regarding vertical refinement.}  The amount of fast memory needed by a running ice sheet simulation is also $O(m)$ if prior states are discarded or transferred to storage.  Ice sheet models also have $O(m)$ velocity variables, but these are not state variables because a very-viscous stress balance will compute velocity as a function of the true state variables.

@@

For each explicit time step the computational cost of a velocity (or velocity/pressure) solution of the stress balance equations is determined by solver design.  We suppose that one such solution requires $O(m^{1+\alpha})$ floating point operations (flops),  with the power $\alpha\ge 0$ depending on the solver implementation.  A Stokes solver using direct linear algebra for each Newton step might yield $\alpha \approx 2$ if sparsity is exploited.  By contrast, a multigrid method \citep{Trottenbergetal2001} will greatly reduce $\alpha$ if correctly applied.  Antarctic ice sheet results by \cite{IsaacStadlerGhattas2015}, for a Stokes solver implemented using algebraic multigrid, show that the number of preconditioned Newton-Krylov iterations \citep{Bueler2021} grows slowly under mesh refinement, suggesting perhaps $\alpha\approx 0.2$ \citep[see Table 8.1]{IsaacStadlerGhattas2015}.
% see data/scaling.m
The algebraic-multigrid based solver in \cite{Lengetal2012} may have similar algorithmic scaling, but reported results do not constrain $\alpha$.  A geometric multigrid method applied to the higher-order stress balance equations \citep{BrownSmithAhmadia2013} suggests $\alpha$ is close to zero for simplified geometries.\footnote{The \cite{BrownSmithAhmadia2013,IsaacStadlerGhattas2015,Lengetal2012} results all show good parallel scaling.}  Note that $\alpha=0$ describes an \emph{optimal} solver  \citep{Bueler2021} in the language of scientific computation.

At the easiest end, the isothermal SIA velocity computation is a trivialization of the Stokes problem in which the vertically-averaged velocity at each node is computed by a pointwise formula.  Therefore, for the SIA a velocity solution requires only $O(m)$ work: $\alpha=0$.

When analyzing solver scaling in our simplified form, one must be aware that the constant in $O(m^{1+\alpha})$ can be very large, and will vary depending on solver design.  In fact, many considerations are suppressed in our simple flops-based analysis of algorithmic scaling, and actual run time is also determined by memory latency, memory bandwidth, and process/thread/GPU parallelism, among other factors.

Regardless of the stress balance source, an explicit time-stepping scheme then applies the mass continuity equation to update the ice thickness using $O(m)$ work.  That is, once the velocity is computed for the geometry from the previous time step, an explicit scheme replaces old thickness values by new ones using a pointwise formula.\footnote{Additional computation needed to remesh the updated geometry, a strongly design-dependent cost, is omitted here.}

\newcommand{\oo}[1]{\displaystyle O\left(#1\right)}
\setlength{\tabcolsep}{5pt}
\renewcommand{\arraystretch}{1.5}
\begin{table*}[ht]
{\normalsize
\begin{tabular}{llll}
\emph{time-stepping} & \emph{dynamics} & \emph{flops per model year} & \emph{[pessimistic stability]} \\ \hline
explicit & SIA    & $\oo{\frac{D\, L^2}{\Delta x^4}} = \oo{\frac{D\, m^2}{L^2}}${\Huge \strut} \\
explicit & Stokes & $\oo{\frac{U L^{2+2\alpha}}{\Delta x^{3+2\alpha}}} = \oo{\frac{U m^{1.5+\alpha}}{L}}${\Huge \strut}\phantom{x} & $\oo{\frac{D\, L^{2+2\alpha}}{\Delta x^{4+2\alpha}}} = \oo{\frac{D\,m^{2+\alpha}}{L^2}}$ \\
implicit & SIA    & $\oo{\frac{q\, L^{2+2\beta}}{\Delta x^{2+2\beta}}} = \oo{q\, m^{1+\beta}}${\Huge \strut} \\
implicit & Stokes & $\oo{\frac{q\, L^{2+2\gamma}}{\Delta x^{2+2\gamma}}} = \oo{q\, m^{1+\gamma}}${\Huge \strut}
\end{tabular}
}
\caption{Asymptotic estimates of algorithmic scaling, measured by floating point operations per model year, for map-plane (2D) time-stepping numerical ice sheet simulations, in the high resolution limit where $\Delta x\to 0$ and $m\to\infty$.  See Table \ref{tab:notation} for notation.}
\label{tab:performancemodel}
\end{table*}

Suppose a numerical model takes time steps of $\Delta t$ model years, equivalently $\Delta t^{-1}$ time steps per model year.  Recalling the scaling of $\Delta x$ with $m$, in \eqref{eq:paramrelation}, we now write time step restrictions as required numbers of time-steps per model year in order to maintain stability, as a function of horizontal resolution $\Delta x$ or degrees of freedom $m$.  In the explicit SIA and pessimistic-Stokes cases we have
\begin{equation}
\frac{1}{\Delta t} > \oo{\frac{D}{\Delta x^2}} = \oo{\frac{D m}{L^2}}. \label{eq:explicitsiarequired}
\end{equation}
The explicit, optimistic-Stokes estimate becomes
\begin{equation}
\frac{1}{\Delta t} > \oo{\frac{U}{\Delta x}} = \oo{\frac{U m^{1/2}}{L}}. \label{eq:explicitoptstokesrequired}
\end{equation}
The number of time steps per model year is then multiplied by the per-step computational cost, namely $O(m^{1+\alpha})$, to give a work estimate for each model year in a simulation.  The results so far are shown in the ``explicit'' rows of Table \ref{tab:performancemodel}.

Unconditionally-stable implicit methods have a fixed time step $\Delta t = 1/q$, independent of $\Delta x$ and determined only by the need to resolve climatic interactions.  On the other hand, the per-step expense is much greater because nontrivial equations, indeed a free-boundary problem, must be solved.  We parameterize the flops of such coupled velocity and geometry-update solutions as $O(m^{1+\beta})$, with the understanding that $\beta \ge \alpha$ for the $\alpha$ value associated to the corresponding fixed-geometry velocity solution.

To our knowledge there is only one implemented, unconditionally-stable, fully-implicit geometric-update solver, namely the SIA scheme in \cite{Bueler2016}.  For simplified dome geometry this method, based on Newton steps solved via direct linear algebra, shows $\beta=1.8$.
% see data/bueler.perf and data/scalingbueler.m
The same work shows how a steady-state ($\Delta t=\infty$) SIA solution is practical for many geometries, but in any case convergence is robust for time steps of years to centuries on kilometer-scale grids, even using realistic, thus irregular, bed topographies.

The hybrid \cite{Brinkerhoffetal2017} flowline solver is implicit but apparently subject to a time-stepping condition when the margin moves, and its algorithmic scaling is unknown.

For an implicit Stokes simulation, a prospective coupled velocity and geometry-update solve is assumed to be, in the absense of constraining research, $O(m^{1+\gamma})$ for some $\gamma>0$ to be determined.  Again $\beta \ge \alpha$ is automatic.  It is not clear if $\gamma\ge \beta$ in general, which might be the expectation, but to the author's knowledge there are no implemented cases to measure.
% The nMCD method proposed in [Bueler 2022] is expensive but multigrid, so it may turn out that $\gamma$ is small.

These comments complete Table \ref{tab:performancemodel}.  Again one must be aware that all asymptotic performance estimates in the Table involve a large, scheme-dependent constant, with this especially true of implicit schemes.

@@

FIXME The real promise of multigrid methods for the implicit, coupled geometry-update plus velocity solution problems; large constants are acceptable if $\beta,\gamma$ are less than 2, for example.  Specifically $\gamma < 1$ might be regarded as the goal of the next decades of research on numerical ice sheet models using Stokes dynamics.  However, this aspiration is long-term because work is barely started on it.
